---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# statgsa

<!-- badges: start -->
<!-- badges: end -->

# statgsa: Automating Epidemiological Data Analysis

*"Anything that can be automated, should be automated."*  
— Inspired by the philosophy of Hadley Wickham's book "R Packages"

Building on this principle, I have developed an entire pipeline for epidemiological data analysis, drawing from years of experience in data science, specifically in biostatistics. The **statgsa** package is designed to streamline and enhance the efficiency of statistical analysis in research, ensuring that critical tasks are automated, consistent, and error-free.

## Why statgsa?

This package offers a comprehensive suite of functions, each meticulously crafted to address a wide array of tasks in statistical analysis. Below are some of the key features:

- **Output Formatting**: Functions designed to ensure that the results of statistical analyses are presented clearly and professionally. This is crucial for organizing research findings in a way that meets the rigorous standards required by scientific journals.
- **Cross Tables with Hypothesis Testing**: Automatically analyzes the data and selects the most appropriate statistical tests, generating detailed cross-tabulations.
- **Numerical Summaries by Group with Hypothesis Testing**: Similar to the cross tables, this feature automatically determines and applies the correct hypothesis tests, providing insightful group summaries.
- **Comprehensive Crosstables**: Generate complete and detailed crosstables that are ready for reporting.
- **Evaluation Metrics for Supervised ML Models**: Includes measures for evaluating the performance of machine learning models in a statistical context.
- **Complex Statistical Analyses**: Streamline and automate intricate statistical procedures, allowing you to focus on interpreting the results.

### The Power of Adaptive Cross Tables

One of the standout features of statgsa is its ability to generate **adaptive cross tables**. These tables automatically adjust based on the types of variables involved, applying the correct hypothesis tests and criteria. This flexibility significantly accelerates the process of generating accurate and publication-ready tables for reports.


## Installation

You can install the development version of statgsa from [GitHub](https://github.com/GabrielGabes/statgsa) with:

```{r carregar o pacote}
# install.packages("pak")
pak::pak("GabrielGabes/statgsa")
```

or

```{r outra forma de carregar o pacote}
# install.packages("remotes")
# remotes::install_github("GabrielGabes/statgsa")
# library(statgsa)
```

## Loading Package

```{r, include = FALSE}
# Suprimir todos os warnings no script
options(warn = -1)
suppressWarnings(library(statgsa))
```

```{r, eval = FALSE}
library(statgsa)
```


Also load dependent packages

```{r pacotes dependentes}
# Package loading facilitator
if(!require(pacman)) install.packages("pacman")
library(pacman)

# dependents Packages
pacman::p_load(
stats,
rlang,
dplyr,
janitor,
effsize,
caret,
DescTools,
car,
pROC,
MuMIn,
magrittr,
tidyr
)
```

## Example

This is a basic example that demonstrates how to solve a common problem in the analysis of epidemiological study data:

The dataset from the `MASS` package contains data on 189 births at the Baystate Medical Centre, Springfield, Massachusetts during 1986 (Venables & Ripley, 2002).

You can find the data dictionary here:  
[Data Dictionary for birthwt](https://www.rdocumentation.org/packages/MASS/versions/7.3-61/topics/birthwt)

### Data Preparation

Let's load the dataset and inspect the first few rows:


```{r base de dados}
library(MASS)
birthwt = MASS::birthwt
head(birthwt)
```

Next, we'll adjust the variable types:

```{r Ajeitando o tipo das variaveis}
for (coluna in names(birthwt)){
  qtd_levels = birthwt[[coluna]] %>% as.factor() %>% levels() %>% length()
  
  if (qtd_levels <= 3){
    birthwt[[coluna]] = birthwt[[coluna]] %>% as.factor()
  }
}
```

## Analysis

The main variable of interest is low birth weight, a binary response variable (D. W. Hosmer & Lemeshow, 1989).

### General Count


```{r contagem simples}
freq_table(birthwt, 'low')
```

### Mother's Race vs. Smoking Habit

```{r tabela cruzada}
count_table(birthwt, 'race', 'smoke')
```

### Mother's Race vs. Mother's Age
We can summarize the age by race using both non-parametric and parametric approaches:

```{r resumo numerico por grupo}
summary_num_nonparametric_groups(birthwt, 'age', 'race')
summary_num_parametric_groups(birthwt, 'age', 'race')
```

### Determining the Appropriate Analytical Approach
Since the t-test assumes normality, we first check the normality of the data:

```{r normalidade por grupo}
group_normality_test(birthwt, 'age', 'race')
```

The third group does not follow a normal distribution, so a non-parametric approach (median [IQR] - Hypothesis Test: Mann-Whitney) is more appropriate. However, you don't need to worry about deciding this manually—just run the code below:

```{r resumo numerico por grupo adaptativo}
summary_num_groups(birthwt, 'age', 'race')
```

### Cross Tables - Univariate Analysis

```{r crosstable}
adaptive_cross_table(birthwt, 'low')
```

### Cross Tables - Changing the Percentage Orientation

```{r crosstable row}
adaptive_cross_table(birthwt, 'low', 'row')
```

### Evaluation Metrics for Supervised ML Models
Here’s how to evaluate a logistic regression model:

```{r modelo logistico}
logit_model <- glm(low ~ age + lwt + race + smoke + ht + ui, data = birthwt, family = binomial)
# summary(logit_model)
binary_model_eval(logit_model) %>% round(2)
univariate_model_analysis(logit_model)
```

And here’s how to evaluate linear regression models:

```{r modelo linear}
model1 <- lm(bwt ~ lwt, data = birthwt)
# summary(model1)
lm_model_eval(model1) %>% round(2)

model2 <- lm(bwt ~ lwt + age + race + smoke + ht + ui, data = birthwt)
# summary(model2)
lm_model_eval(model2) %>% round(2)
```
